---
title: "STDM ARIMA Report"
output: html_document
date: "2025-03-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown for ARIMA Modelling

#Loading Packages and Data Preprocessing

#Read excel file, install packages if unavailable
```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(lubridate)
library(imputeTS)
library(ggplot2)
library(forecast)
library(spdep)
library(tseries)
library(sf)
library(spData)
library(spatialreg)
```

# Reading Excel sheets separately as spatial components are in different sheets
``` {r}
data <- read_excel("Design-Hour-Report-2023.xlsx", sheet = "Top 200 Hours by Site", skip = 3)
site_coordinates <- read_excel("Design-Hour-Report-2023.xlsx", sheet = "Site-Specific Statistics", skip = 4)
```

#Rename columns
``` {r}
colnames(data) <- c("Site_Number", "Year", NA, NA, "Count_Date", "Day_of_Week", "Hour", "Volume_Rank", "Traffic_Volume", "Decreasing_Dir_Volume", "K_Factor", "D_Factor", NA, NA)
```

# Select relevant columns and remove unnecessary columns
```{r}
data <- data %>% 
  select(Site_Number, Year, Count_Date, Day_of_Week, Hour, Volume_Rank, Traffic_Volume, K_Factor, D_Factor) %>%
  drop_na(Site_Number, Count_Date, Traffic_Volume) #Drop rows with missing values
```

#Convert to POSIXct format on the date and time columns
```{r}
data <- data %>%
  mutate(Count_Date = as.Date(Count_Date),
         Traffic_Volume = as.numeric(Traffic_Volume),
         DateTime = as.POSIXct(paste(Count_Date, Hour), format = "%Y-%m-%d %H")
         )
```

#Filter data for first station and group by monthly intervals
```{r}
data_selected <- data %>%
  filter(Site_Number == unique(Site_Number) [1]) %>%
  mutate(Month = floor_date(DateTime, "month")) %>%
  arrange(DateTime)
```

#Handle missing values with interpolation
```{r}
data_selected$Traffic_Volume <- na_interpolation(data_selected$Traffic_Volume)
```

#Check for missing values
```{r}
sum(is.na(data_selected$Traffic_Volume))
```

#Apply interpolation to Traffic Volume
```{r}
data_selected$Traffic_Volume <- na_interpolation(data_selected$Traffic_Volume)
```

#Time Series Plot - ESDA for one site
```{r}
ggplot(data_selected, aes(x = DateTime, y = Traffic_Volume)) + geom_line(color = "blue") +
  labs(title = "Traffic Volume Over Time",
       x = "Date", y = "Traffic Volume") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# KDE as ESDA for Spatial Component, clean map location to separate Lat long first

```{r}
coords_clean <- site_coordinates %>%
  select("Site", "Map Location") %>%
  separate("Map Location", into = c("Latitude", "Longitude"), sep = ",") %>%
  mutate(Latitude = as.numeric(Latitude), Longitude = as.numeric(Longitude))
```

```{r}
data_selected_kde <- data %>%
  mutate(Month = floor_date(DateTime, "month")) %>%
  arrange(DateTime)
```

#Ensure no NA values
```{r}
data_selected_kde$Traffic_Volume <- na_interpolation(data_selected_kde$Traffic_Volume)
```

#Merging
```{r}
data_with_coords <- data_selected_kde %>%
  filter(!is.na(Traffic_Volume)) %>%
  left_join(coords_clean, by = c("Site_Number" = "Site"))
```

#Compute the Traffic_Volume_Change column (difference in Traffic Volume)
```{r}
data_with_coords <- data_with_coords %>%
  arrange(Site_Number, DateTime) %>%
  group_by(Site_Number) %>%
  mutate(Traffic_Volume_Change = Traffic_Volume - lag(Traffic_Volume)) %>%
  ungroup()
```

#Filter out NA values
```{r}
data_with_coords <- data_with_coords %>%
  filter(!is.na(Traffic_Volume_Change))
```

#KDE plot for all sites
```{r}
ggplot(data_with_coords, aes(x = Longitude, y = Latitude)) +
  geom_density_2d(aes(color = Traffic_Volume_Change)) + scale_color_viridis_c() + labs(title = "Density Estimation of Traffic Volume Change") +
  theme_minimal()
```

# ACF and PACF for ARIMA in one site
# One site is chosen due to the complex nature of the original dataset
```{r}
acf(data_selected$Traffic_Volume, main = "ACF ARIMA", lag.max = 36)
```
```{r}
pacf(data_selected$Traffic_Volume, main = "PACF ARIMA", lag.max = 36)
```
#Stationarity Check using Augmented Dickey-Fuller (ADF) test
```{r}
adf_test <- adf.test(data_selected$Traffic_Volume, alternative = "stationary")

print(adf_test)
```

# ---ARIMA Model---
```{r}
#Estimate parameters and fit ARIMA model with seasonal components
arima_model <- arima(data_selected$Traffic_Volume, order = c(1,0,1), seasonal = list(order = c(2,0,2), period = 12))

summary(arima_model)
```

#Diagnostic Checking
```{r}
acf(residuals(arima_model))
```

```{r}
plot(residuals(arima_model))
```

```{r}
#Test using ljung-box test
ljung_box_test <- Box.test(residuals(arima_model), lag = 12, type = "Ljung-Box")
print(ljung_box_test)
```

# p value is more than 0.05 indicating good fit

#Forecast/Predict next 12 values (one step ahead)
```{r}
arima_forecast <- predict(arima_model, n.ahead = 12)
```

#Plot observed vs forecasted (one step ahead) for ARIMA prediction
```{r}
# Select the last 12 observed values
last_12_obs <- tail(data_selected$Traffic_Volume, 12)

# Time index for plotting (aligning forecast with observed values)
time_index <- seq(length(last_12_obs) + 1,
                  length(last_12_obs) + 12)

matplot(c(1:12), cbind(last_12_obs, arima_forecast$pred),
          type = "l",
          col = c("black", "red"), lty = c(1,2),
          xlab = "Time", ylab = "Traffic Volume",
          main = "ARIMA: Observed vs Forecasted")
```

#Residual Check
```{r}
# Check for autocorrelation in residuals
checkresiduals(arima_model)

# Perform Ljung-Box test
Box.test(residuals(arima_model), lag = 12, type = "Ljung-Box")
```

#RMSE for ARIMA
```{r}
fitted_values <- fitted(arima_model)

actual_values <- data_selected$Traffic_Volume[(length(data_selected$Traffic_Volume) - length(fitted_values) + 1): length(data_selected$Traffic_Volume)]

#Calculate RMSE
arima_rmse <- sqrt(mean((fitted_values - actual_values)^2, na.rm = TRUE))
print(paste("ARIMA RMSE:", arima_rmse))
```

#ARIMA model Fitting
```{r}
#using Arima from forecast package
fit_Ar <- Arima(data_selected$Traffic_Volume[1:(nrow(data_selected)-12)],
                order = c(1,0,1), seasonal = list(order = c(2,0,2), period = 12))

pre_Ar <- Arima(data_selected$Traffic_Volume[(nrow(data_selected)-11):nrow(data_selected)],
                model = fit_Ar)

matplot(cbind(pre_Ar$fitted, pre_Ar$x), type = "l", col = c("red", "black"), lty = c(1,2),
        xlab = "Time", ylab = "Traffic Volume", main = "ARIMA Predictions")
```

#Auto ARIMA comparison
```{r}
auto_arima_model <- auto.arima(data_selected$Traffic_Volume)

checkresiduals(auto_arima_model)

# Ljung-Box Test
Box.test(residuals(auto_arima_model), lag = 12, type = "Ljung-Box")
```

#Model Fitting with Auto ARIMA
```{r}
auto_arima_forecast <- forecast(auto_arima_model, n.ahead = 12)
plot(auto_arima_forecast)
```

#RMSE for Auto ARIMA
```{r}
auto_fitted_values <- fitted(auto_arima_model)

actual_value_auto <- data_selected$Traffic_Volume[(length(data_selected$Traffic_Volume) - length(auto_fitted_values) + 1):length(data_selected$Traffic_Volume)]

#Calculate RMSE
auto_arima_rmse <- sqrt(mean((auto_fitted_values - actual_value_auto)^2, na.rm = TRUE))
print(paste("Auto ARIMA RMSE:", auto_arima_rmse))
```

#Code References for ARIMA modelling and deciding p,d,q values: 
#https://otexts.com/fpp2/arima-r.html
#https://www.geo.fu-berlin.de/en/v/soga-r/Advances-statistics/Time-series-analysis/ARIMA-modelling-in-R/index.html
#https://moodle.ucl.ac.uk/pluginfile.php/8470339/mod_resource/content/24/_book/statistical-modelling-of-time-series-and-spatio-temporal-series.html#the-arima-modelling-framework 
